---
title: "00_all"
format: html
editor: visual
---

# Run complete project

By running this quarto document, the entire project will be run. Note: Initially we tried to use quarto::quarto_render() but this function would not let us save the generated files in a different location, there is an [issue](https://github.com/quarto-dev/quarto-cli/discussions/2171) in github and apparently could not be resolved. That is the reason why we used rmarkdown::render instead.

```{r}
library("quarto")
library("here")
```

```{r}
source("99_proj_func.R")

```

## Import + Tidy

### Download and write data

```{r}
rmarkdown::render(here("R/01_download_data.qmd"), 
                  output_dir = here("results"))

```

## Cleaning

```{r}
rmarkdown::render(here("R/02_clean_files.qmd"), 
                  output_dir = here("results"))
```

## Transform - Visualize - Model

```{r}
rmarkdown::render(here("R/03_description.qmd"), 
                  output_dir = here("results"))
```

```{r}
rmarkdown::render(here("R/04_analysis_1.qmd"), 
                  output_dir = here("results"))

```

```{r}

rmarkdown::render(here("R/05_analysis_2.qmd"), 
                  output_dir = here("results"))

```

```{r}
quarto::quarto_render(
  input = here("doc","presentation.qmd"))
```
